---
title: "Banding processing"
author: "LGCarlson"
date: "9/15/2020"
output: html_document
---

## Banding codes

_age:_
0 unknown
1 AHY
2 HY
3 Juvenile
4 Local
5 SY
6 ASY

_sex:_
0 unknown
4 M
5 F

_b_dir:_
0 NW quarter of the earth

_b_flyway:_
2 Mississippi
3 Central
6 Canada

_b_region:_
[codes here](https://www.pwrc.usgs.gov/BBL/MANUAL/reg.cfm#region)
Canadian categories to keep: (US Flyways okay)
43 Northwest Territories
40 Nunavut
04 Alberta
79 Saskatchewan
45 Manitoba
68 Ontario

_b_day_code:_
0 unknown
1 Day 1-5
2 Day 6-10
3 Day 11-15
4 Day 16-20
5 Day 21-25
6 Day >26

_status:_
2 transported
3 normal wild
4 hand reared
5 sick, injured, deformity
6,7,8 experimental, etc

_sp_num:_
1320 MALL

_ai_vai:_
Categories to keep:
Federal numbered metal band only. 
Captured by spotlighting. 
Misc. metal band (State, Provincial etc) with address or telephone number, plus Federal band.
Miscellaneous band, Federal band, plus auxiliary marker(s). 

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(janitor)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(sf)
```

```{r}
theme_set(theme_light())

#set personalized plotting theme
LGCtheme <- theme_light() + theme(panel.background = element_blank(), 
                panel.grid = element_blank(), 
                axis.text = element_text(size = 13), 
                axis.title = element_text(size = 14), 
                legend.text = element_text(size = 13), 
                legend.title = element_text(size = 14), 
                strip.background = element_blank(), 
                strip.text = element_text(size = 13, color = "black"))

gmri_qual_extended<-c("#003A54","#00608A","#5D96AF","#3F4B24","#6F7E4B","#97AA5C","#A8B212","#AF9700","#EACA00","#FFE961","#BB3F0E", "#EA4F12","#F4916B","#00736D", "#55A29E","#A0CDCB","#767676","#A7A7A7","#E2E2E2")
```


```{r}
date_code_tib<-tribble(
  ~b_day_code, ~b_day, 
  1, 5, 
  2, 10, 
  3, 15, 
  4, 20, 
  5, 25, 
  6, 28)


provs_keep<-c(43,40,04,79,45,68)

ai_vai_vec<-c("Federal numbered metal band only.","Captured by spotlighting.", "Misc. metal band (State, Provincial etc) with address or telephone number, plus Federal band.","Miscellaneous band, Federal band, plus auxiliary marker(s).")
```

## Load data

```{r, echo=FALSE, message=F}
ca_band<-read_csv(here::here("raw_data", "canada_banding.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(b_flyway_name = "Canada") %>% 
  filter(b_region %in% provs_keep)

cl_band<-read_csv(here::here("raw_data", "central_banding.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(b_flyway_name = "Central")

ms_band<-read_csv(here::here("raw_data", "miss_banding.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(b_flyway_name = "Mississippi")
```

_Modifications:_

* remove indv. where banding month is unknown

* remove indv. where banding day is unknown

* join date_code_tib by b_day_code, so that b_day so that approximate banding_dates can be created using maximum dates in b_day_code range

* create a new column with banding date


```{r}
all_bands<-bind_rows(ca_band, cl_band, ms_band) %>% 
  filter(b_month <= 12, 
         b_day_code > 0) %>% 
  left_join(date_code_tib, by = "b_day_code") %>% 
  mutate(b_date = as.Date(paste(b_year,b_month,b_day, sep = "-"))) 
```

## Check data

```{r}
table(all_bands$age)
table(all_bands$sex)
table(all_bands$b_month)
table(all_bands$b_day_code)
table(all_bands$status)
#table(all_bands$ai_vai)
```

## Processing code

_Modifications:_

* limit years to 1961-2011

* keep only pre-season banded birds

* keep only normal, wild birds (status)

* keep only known sexes (4,5)

* keep only desired ai_vai's

* re-order columns manually

```{r}
final_bands<-all_bands %>% 
  filter(b_year >=1961 & b_year <=2011) %>% 
  filter(b_month >=6 & b_month <=9) %>% 
  filter(status == 3) %>% 
  filter(sex == 4 | sex == 5) %>% 
  filter(ai_vai %in% ai_vai_vec) %>% 
  dplyr::select(band_prefix_plus,count_of_birds,age,age_vage,sex,sp_num,status,b_flyway,b_flyway_name,b_date,b_year,b_month,b_day_code, b_day, country_code,b_region,state_code,b_dir,b_coord_prec,b_lat,b_long,gisb_lat,gisb_long, band_size,band_type,ai_vai,how_aged,how_sexed)
```

## Visualize data

```{r, fig.height=4, fig.width=10}
final_bands %>% 
  group_by(b_year, sex, b_flyway_name) %>% 
  summarise(total_banded = sum(count_of_birds)) %>% 
  ggplot(aes(x=b_year, y=total_banded, color=as.factor(sex))) + geom_line() + 
  facet_wrap(~b_flyway_name, ncol=3) + 
  scale_color_manual(values=c("#5D96AF","#C07F9C")) + LGCtheme + theme(legend.position = "none") +
  labs(x="Banding Year", y="Total banded", title="MALLs banded in each flyway, by banding year, by sex")
```




```{r, fig.height=4.5, fig.width=7.5}
final_bands %>% 
  mutate(b_month = as.factor(b_month)) %>% 
  group_by(b_year, b_month) %>% 
  summarise(total_banded = sum(count_of_birds)) %>% 
  ggplot(aes(x=b_year, y=total_banded, color=forcats::fct_reorder(b_month, total_banded, .desc = T))) + 
  geom_line() + scale_color_manual(values = c(gmri_qual_extended[3],gmri_qual_extended[6],
                                              gmri_qual_extended[13],gmri_qual_extended[16])) + LGCtheme +
  labs(x="Banding Year", y="Total banded", 
       title="MALLs banded by month", color="Banding month")
```

## Spatial visualization

```{r, message=F, warning=F, echo=F}
# load in world and state sf
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

states <- rnaturalearth::ne_states(country = c("canada", "united states of america"), returnclass = "sf")


#load in flyway shapefiles
cent_flyway <- sf::st_read(here::here("shapefiles","Central_Flyway.shp")) %>% 
  st_transform(crs = "+init=epsg:4326") 

miss_flyway <- sf::st_read(here::here("shapefiles","Mississippi_Flyway.shp")) %>% 
  st_transform(crs = 4326) 
```

```{r}
banding_locs<-final_bands %>% 
  group_by(gisb_lat,gisb_long) %>% 
  summarise(n_bands=n())
```


```{r, fig.height=8, fig.width=8}
ggplot(data = world) + geom_sf(fill="white") + 
  geom_sf(data=states, fill=NA) + 
  geom_sf(data=cent_flyway, fill="grey") +
  geom_sf(data=miss_flyway, fill="lightgrey") + 
  geom_point(data=banding_locs, aes(x=gisb_long, y=gisb_lat, size=n_bands)) +
  coord_sf(xlim = c(-130, -70), ylim = c(25, 70), expand = FALSE) + 
  labs(x="Banding longitude", y="Banding latitude", size="Number banded")
```


## Save datafile

```{r}
## write_csv(final_bands, here::here("proc_data","banding_data.csv"))
```

